Pipit on the Post: Proving Pre- and Post-conditions of Reactive Systems

### 2023-12-19 2024pipit/Main.tex
ok

@@

Big TODOs:
\begin{itemize}
  \item example: trigger-fetch from ttcan;
  \item transition systems and translation;
  \item C codegen: note about duplicating work, sketch potential solution;
  \item C codegen: future work: sketch potential solution for separate compilation, compare to Accelerate FFI \cite{clifton2014embedding};
  \item cite Liam \cite{oconnor2019deferring} similar idea, but we defer the \emph{context} rather than just syntactically deferring the proof;
  \item oracle transition system in the semantics: does this have a standard name? references? not really related to prophecy variables.
  % \item 
\end{itemize}

>>>>

\ifx\authoranonymous\relax
\newcommand{\GITHUBURL}{(anonymised)}
\else
\newcommand{\GITHUBURL}{\url{https://github.com/songlarknet/pipit}}
\fi

@@

Big TODOs:
\begin{itemize}
  \item example: trigger-fetch from ttcan;
  \item transition systems and translation;
\end{itemize}

### 2023-12-19 2024pipit/section/01-Introduction.tex
ok

%!TEX root = ../Main.tex
\section{Introduction}
Safety-critical control systems, such as the anti-lock braking systems that are present in most cars today, need to be correct and execute in real-time.
One approach, favoured by parts of the aerospace industry, is to implement the controllers in a high-level language such as Lustre~\cite{caspi1995functional} or Scade~\cite{colaco2017scade}, and verify that the implementations satisfy the high-level specification using a model-checker, such as Kind2~\cite{champion2016kind2}.
These model-checkers can prove many interesting properties automatically, but do not provide many options for manual proofs when the automated proof techniques fail.
Additionally, the semantics used by the model-checker may not match the semantics of the compiled code, in which case properties proved do not necessarily hold on the real system.
This mismatch may occur even when the compiler has been verified to be correct, as in the case of Vélus~\cite{bourke2017formally}.
For example, in Vélus, integer division rounds towards zero, matching the semantics of C; however, integer division in Kind2 rounds to negative infinity, matching SMT-lib~\cite{BarFT2016SMTLIB,kind2023intdiv}.

To be confident that our proofs hold on the real system, we need a single semantics that is shared between the compiler and the model-checker or prover.
In this paper we introduce Pipit\footnote{Implementation available at \url{https://github.com/songlarknet/pipit}}, an embedded domain-specific language for implementing and verifying controllers in \fstar{}.
Pipit aims to provide a high-level language based on Lustre, while reusing \fstar{}'s proof automation and manual proofs for verifying controllers~\cite{martinez2019meta}, and using \lowstar{}'s C-code generation for real-time execution~\cite{protzenko2017verified}.
% By designing Pipit as an embedded language, we can reuse many of the \fstar{} meta-language's features while retaining a small core language.
Pipit translates its expression language to a transition system for k-inductive proofs, which is verified; verifying the translation to imperative code is ongoing.

>>>>

%!TEX root = ../Main.tex
\section{Introduction}
Safety-critical control systems, such as the anti-lock braking systems that are present in most cars today, need to be correct and execute in real-time.
One approach, favoured by parts of the aerospace industry, is to implement the controllers in a high-level language such as Lustre~\cite{caspi1995functional} or Scade~\cite{colaco2017scade}, and verify that the implementations satisfy the high-level specification using a model-checker, such as Kind2~\cite{champion2016kind2}.
These model-checkers can prove many interesting safety properties automatically, but do not provide many options for manual proofs when the automated proof techniques fail.
Additionally, the semantics used by the model-checker may not match the semantics of the compiled code, in which case properties proved do not necessarily hold on the real system.
This mismatch may occur even when the compiler has been verified to be correct, as in the case of Vélus~\cite{bourke2017formally}.
For example, in Vélus, integer division rounds towards zero, matching the semantics of C; however, integer division in Kind2 rounds to negative infinity, matching SMT-lib~\cite{BarFT2016SMTLIB,kind2023intdiv}.

To be confident that our proofs hold on the real system, we need a single semantics that is shared between the compiler and the model-checker or prover.
In this paper we introduce Pipit\footnote{Implementation available at \GITHUBURL}, an embedded domain-specific language for implementing and verifying controllers in \fstar{}.
Pipit aims to provide a high-level language based on Lustre, while reusing \fstar{}'s proof automation and manual proofs for verifying controllers~\cite{martinez2019meta}, and using \lowstar{}'s C-code generation for real-time execution~\cite{protzenko2017verified}.
% By designing Pipit as an embedded language, we can reuse many of the \fstar{} meta-language's features while retaining a small core language.
Pipit translates its expression language to a transition system for k-inductive proofs, which is verified; verifying the translation to imperative code is ongoing.

### 2023-12-19 2024pipit/section/70-Related.tex
ok

>>>>

The deferred aspect of our proofs is similar to the deferred proofs of verification conditions for imperative programs, such as \cite{oconnor2019deferring}.
However, such verification conditions are \emph{syntactically} deferred so that the verification condition can be proved later; in our case, the verification conditions are \emph{semantically} deferred, so that more knowledge of the enclosing program can be exploited in the proof.
In imperative programs, this sort of extra knowledge is generally provided explicitly as loop invariants, and non-looping statements have their weakest precondition computed automatically.
In Lustre-style reactive languages such as ours, programs tend to be composed of many nested recursive streams, which perform a similar function to loops.
Explicitly specifying an invariant for each recursive stream would be cumbersome; deferring the proof allows such invariants to be implicit.

### 2023-12-19 2024pipit/section/80-Conclusion.tex
ok

Our preliminary results show that \fstar{}'s proof automation and code extraction are suitable for verifying reactive systems and executing them in real-time; these results still require further work.
Next, we intend to verify the imperative code generation.
To verify large programs, we also need some way to separately prove smaller pieces which can then be composed together, such as contracts~\cite{champion2016kind2}.
Finally, we need to evaluate Pipit on larger control systems before extending the language to support more features, such as Lustre's clocks for describing partially-defined streams~\cite{caspi1995functional}.

@@

>>>>

Our preliminary results show that \fstar{}'s proof automation and code extraction are suitable for verifying reactive systems and executing them in real-time; these results still require further work.
Next, we intend to verify the imperative code generation.
%  To verify large programs, we also need some way to separately prove smaller pieces which can then be composed together, such as contracts~\cite{champion2016kind2}.
Finally, we need to evaluate Pipit on larger control systems before extending the language to support more features, such as Lustre's clocks for describing partially-defined streams~\cite{caspi1995functional}.

@@

We are interested in further pursuing the intersection of model-checking with interactive theorem proving.
A smart contract called Djed \cite{zahnentferner2023djed} currently uses a mixture of Kind2 \cite{champion2016kind2} and manual Isabelle/HOL proofs to show that the contract is well-behaved.
In future work, we would like to further investigate whether Pipit's integration of streaming proofs with \fstar{}'s automated proof system would be able to provide similar proofs, without introducing any semantic gap between the two systems.

Our current array support is limited: constant arrays provide a pure index function, while we only support fixed-size mutable arrays by wrapping bit-vectors.
Array support is an obvious direction for future work; integrating with a verified array-fusion system such as \cite{robinson2017machine} would be an interesting and useful extension.

Future work: verified imperative code generation.
